---
title: "Cumulative WAR"
author: "Brice Keown"
date: "`r Sys.Date()`"
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'index.html'))})
---

##Cumulative WAR
I was playing around with the Lahman data, and got curious about cumulative player WAR by team.  What are the top N players by position for each team?



```{r Setup, message=FALSE}

# package load 
library(Lahman)
library(ggplot2)
library(dplyr)
library(tidyverse)
library(ggrepel)
library(gghighlight)
library(here)
library(lattice)
library(directlabels)

```

#### Read and summarize the data


```{r, message=FALSE}
data(Teams)
data("People")
data("HallOfFame")
people_tb <- as_tibble(People)
people_tb
#Get WAR
WAR_data <- read_csv(file = "https://www.baseball-reference.com/data/war_daily_pitch.txt")  #%>% filter(name_common=="Ted Williams")
WAR_tb <- as_tibble(WAR_data)
WAR_tb
#hofinfo
```



```{r, message=FALSE}
```

Adds player names to HOF, only gets folks who're inducted, join to get name_common
```{r, message=FALSE}
HOF <- HallOfFame %>%
  filter(inducted == "Y", category =="Player")

HOF_with_names <- HOF %>% inner_join(People, by = "playerID") %>% unite("name_common","nameFirst":"nameLast",sep=" ", remove=TRUE)
HOF_with_names_tb <- as_tibble(HOF_with_names)
#HOF_with_names_tb

#WAR_tb <- as_tibble(WAR_data)
#WAR_tb
colnames(WAR_data)[colnames(WAR_data) == 'player_ID'] <- 'bbrefID'

HOF_with_names_and_war <- HOF_with_names %>% inner_join(WAR_data, by="bbrefID") 
HOF_with_names_and_war$WAR <- as.numeric(HOF_with_names_and_war$WAR)


HOF_with_names_and_war_summed <- HOF_with_names_and_war %>% 
  group_by(name_common.x, yearID) %>% 
  summarise(WAR = sum(WAR)) %>% filter(WAR>5 )

p <- ggplot(HOF_with_names_and_war_summed, mapping=aes(x=yearID,y=WAR,label=name_common.x))
p + geom_point() + geom_text_repel(aes(label=name_common.x),hjust=0, vjust=0)

HOF_with_names_and_war_summed <- as_tibble(HOF_with_names_and_war_summed)
HOF_with_names_and_war_summed
```




```{r}
HOF_with_names_and_war

HOF_with_names_and_war_cumsummed <- HOF_with_names_and_war %>% 
  group_by(name_common.x, year_ID) %>% summarise(WAR = sum(WAR), age=age) %>% 
  group_by(name_common.x) %>% summarise(cumWAR = cumsum(WAR), year_ID, age, sumWAR = sum(WAR)) %>% filter(sumWAR >25) 
HOF_with_names_and_war_cumsummed


#HOF_with_names_and_war_cumsummed_unfuck <- HOF_with_names_and_war_cumsummed %>% group_by(name_common, year_ID) %>% summarise(cumWAR= sum(cumWAR),sumWAR)
#HOF_with_names_and_war_cumsummed_unfuck


p <- ggplot(HOF_with_names_and_war_cumsummed, mapping=aes(x=year_ID,y=cumWAR))
p + geom_line(aes(group=name_common.x, colour=name_common.x), alpha = .7) + gghighlight(max(cumWAR)>100)+
    labs(x="Year", y = "Cumulative WAR", 
       title = "Cumulative WAR for Hall of Fame Pitchers",
       subtitle = "Career WAR > 100 highlighted",
       caption="Source: Baseball Reference")
p + geom_line(aes(group=name_common.x, colour=name_common.x), alpha = .7) + gghighlight(max(cumWAR)<40)+
    labs(x="Year", y = "Cumulative WAR", 
       title = "Cumulative WAR for Hall of Fame Pitchers",
       subtitle = "Career WAR < 40 highlighted",
       caption="Source: Baseball Reference")

p <- ggplot(HOF_with_names_and_war_cumsummed, mapping=aes(x=age,y=cumWAR))
p + geom_line(aes(group=name_common.x), alpha=.1)

#HOF_with_names_and_war_cumsummed <- as_tibble(HOF_with_names_and_war_cumsummed)
#HOF_with_names_and_war_cumsummed
```
OK so moving on to trying to do small multiples of teams.  Feeling good about all that though

```{r}
WAR_data
WAR_data$WAR <- as.numeric(WAR_data$WAR)
#Fix Team Names, need to figure out how to get unique
#Team_Map <- Teams %>% group_by(teamID) %>% summarise(franchID, yearID) %>% distinct()
#Team_Map

#WAR_teamfix <- merge(x=WAR, y=Team_Map, by.x = c("team_ID","year_ID"), by.y= c("teamID","yearID") )
##WAR_teamfix

WAR_teamfix <- WAR_data %>% mutate(team_ID=recode(team_ID,
                                                  "ANA" = "LAA",
                                                  "CAL" = "LAA",
                                                  "TBA" = "TBD",
                                                  "TBR" = "TBD"))

WAR_cum <- WAR_teamfix %>% group_by(bbrefID)  %>% summarise(cumWAR =cumsum(WAR), maxWAR=sum(WAR),name_common, WAR, year_ID, lg_ID, age, team_ID) %>% group_by(bbrefID, team_ID) %>% summarise(cumWAR_by_team = cumsum(WAR), maxWAR_by_team = sum(WAR),maxYear_by_team = max(year_ID),   name_common, WAR, year_ID, lg_ID, age, team_ID)

#need to fix team names!


```
making sure just the mariners works
```{r}
WAR_cum_justms <- WAR_cum %>% filter(team_ID == "SEA", year_ID > 1900)
WAR_cum_justms_top5 <- WAR_cum_justms %>%slice_max(order_by=cumWAR_by_team,n=1) %>% ungroup() %>% slice_max(order_by=cumWAR_by_team,n=5) %>% summarise(bbrefID, top5=TRUE, maxYear_by_team)
WAR_cum_justms_top5

war_cum_justmswithtop5 <- left_join(WAR_cum_justms,WAR_cum_justms_top5)
war_cum_justmswithtop5

WAR_cum_justms_top5
p <- ggplot(war_cum_justmswithtop5, mapping=aes(x = year_ID, y=cumWAR_by_team))
p + geom_line(aes(group=bbrefID), alpha = .3) + geom_line(subset(war_cum_justmswithtop5, top5==TRUE), mapping=aes(group=bbrefID, colour = name_common)) + geom_label_repel(subset(war_cum_justmswithtop5, top5==TRUE & maxYear_by_team==year_ID), mapping=aes(group=bbrefID, colour = name_common, label = name_common),point.padding = NA, nudge_y = .5 )
# p + geom_line(subset(war_cum_justmswithtop5, top5==TRUE), mapping=aes(group=bbrefID, label))
```
now lets do 2 teams
```{r}
WAR_cum_just2 <- WAR_cum %>% filter(team_ID == "NYY" | team_ID =="CHC") %>% filter(year_ID > 1960)
TopN <- 5
WAR_cum_just2_topn <- WAR_cum_just2 %>%slice_max(order_by=cumWAR_by_team,n=1) %>% ungroup() %>% group_by(team_ID) %>% slice_max(order_by=cumWAR_by_team,n=TopN) %>% summarise(bbrefID, team_ID, isTopN=TRUE)
WAR_cum_just2_topn


war_cum_just2withtopn <- left_join(WAR_cum_just2,WAR_cum_just2_topn)
war_cum_just2withtopn[is.na(war_cum_just2withtopn)] <- 0
war_cum_just2withtopn


p <- ggplot(war_cum_just2withtopn, mapping=aes(x = year_ID, y=cumWAR_by_team))
p + geom_line(aes(group=bbrefID), alpha = .2) + geom_line(subset(war_cum_just2withtopn, isTopN==TRUE), mapping=aes(group=bbrefID, colour = name_common)) +
    geom_label_repel(subset(war_cum_just2withtopn, isTopN==TRUE & cumWAR_by_team==maxWAR_by_team), mapping=aes(group=bbrefID, colour = name_common, label = name_common),point.padding = NA, nudge_y = 0, force = 3, max.iter=10000,fill = alpha(c("white"),0.4)) + facet_wrap(~team_ID)
```

try all teams!
```{r fig.width=1, fig.height=8}

MLB <- data.frame("League" = c("AL","AL","AL","AL","AL","AL","AL","AL","AL","AL","AL","AL","AL","AL","AL","NL","NL","NL","NL","NL","NL","NL","NL","NL","NL","NL","NL","NL","NL","NL"),"Division"=c("West","West","West","West","West","Central","Central","Central","Central","Central","East","East","East","East","East","West","West","West","West","West","Central","Central","Central","Central","Central","East","East","East","East","East"), "Team"=c("SEA","OAK","LAA", "TEX", "HOU","CLE","CHW","MIN","DET","KCR","NYY","BOS","TOR","BAL","TBD","SDP","LAD","COL","ARI","SFG","CIN","CHC","STL","MIL","PIT","NYM","PHI","WSN","FLA", "ATL"))

WAR_cum_all <- left_join(WAR_cum, MLB, by = c("team_ID" = "Team")) %>% unite("Division","League":"Division",sep=" ", remove=TRUE)
WAR_cum_all
ThisLeague <- "AL Central"
#WAR_cum_all <- WAR_cum %>% filter(team_ID == "SEA" | team_ID =="OAK" | team_ID == "LAA"| team_ID=="TEX" | team_ID == "HOU", year_ID > 1900)
WAR_cum_all <- WAR_cum_all %>% filter( year_ID > 1960 & Division != "NA NA")
WAR_cum_all
TopN <- 5
WAR_cum_all_topn <- WAR_cum_all %>%slice_max(order_by=cumWAR_by_team,n=1) %>% ungroup() %>% group_by(team_ID) %>% slice_max(order_by=cumWAR_by_team,n=TopN) %>% summarise(bbrefID, team_ID, Division, isTopN=TRUE)

WAR_cum_all_topn <- left_join(WAR_cum_all,WAR_cum_all_topn)
#WAR_cum_all_topn[is.na(WAR_cum_all_topn)] <- 0
WAR_cum_all_topn

p <- ggplot(WAR_cum_all_topn, mapping=aes(x = year_ID, y=cumWAR_by_team))
p_out <- p + geom_line(aes(group=bbrefID), alpha = .1) + 
  geom_line(subset(WAR_cum_all_topn, isTopN==TRUE), mapping=aes(group=bbrefID, colour = name_common)) +
  geom_label_repel(subset(WAR_cum_all_topn, isTopN==TRUE & cumWAR_by_team==maxWAR_by_team),
                   mapping=aes(group=bbrefID, colour = name_common, label = name_common),point.padding = NA, nudge_y = .9, force = 3, max.iter=50000,fill = alpha(c("white"),0.4) ) + 
  facet_wrap(~team_ID, ncol=5) + theme(legend.position = "none",panel.spacing = unit(1, "lines")) +
  labs(x="Year", y = "Cumulative WAR", 
       title = paste("Cumulative WAR for Pitchers by Team, 1960-2019"),
       subtitle = "Top 5 player-team-WAR highlighted",
       caption="Source: Baseball Reference")
ggsave(here(paste("docs/All30since1960_labels.png")),plot=p_out, height =18, width = 15, units ="in")
                                                     

```
WAR_cum_all_topn <- left_join(WAR_cum_all,WAR_cum_all_topn)
#WAR_cum_all_topn[is.na(WAR_cum_all_topn)] <- 0
WAR_cum_all_topn

p <- ggplot(WAR_cum_all_topn, mapping=aes(x = year_ID, y=cumWAR_by_team))
p_out <- p + geom_line(aes(group=bbrefID), alpha = .3) + 
  geom_line(subset(WAR_cum_all_topn, isTopN==TRUE), mapping=aes(group=bbrefID, colour = name_common)) +
  geom_text_repel(subset(WAR_cum_all_topn, isTopN==TRUE & cumWAR_by_team==maxWAR_by_team),
                   mapping=aes(group=bbrefID, colour = name_common, label = name_common)) + 
  facet_wrap(~team_ID, ncol=5,scales = "free_x") + theme(legend.position = "none",panel.spacing = unit(1, "lines")) +
  labs(x="Year", y = "Cumulative WAR", 
       title = paste("Cumulative WAR for Pitchers by Team, " , ThisLeague),
       subtitle = "Top 5 player-team-WAR highlighted",
       caption="Source: Baseball Reference")
ggsave(here(paste(ThisLeague, ".png")),plot=p_out, height =4, width = 16, units ="in")


p <- ggplot(WAR_cum_all_topn, mapping=aes(x = year_ID, y=cumWAR_by_team))
p_out <- p + geom_line(aes(group=bbrefID), alpha = .3) + 
  geom_line(subset(WAR_cum_all_topn, isTopN==TRUE), mapping=aes(group=bbrefID, colour = name_common)) +
  geom_text_repel(subset(WAR_cum_all_topn, isTopN==TRUE & cumWAR_by_team==maxWAR_by_team),
                   mapping=aes(group=bbrefID, colour = name_common, label = name_common)) + 
  facet_grid(rows = vars(Division), col = vars(team_ID), drop = TRUE) + theme(legend.position = "none",panel.spacing = unit(1, "lines")) +
  labs(x="Year", y = "Cumulative WAR", 
       title = paste("Cumulative WAR for Pitchers by Team, " , ThisLeague),
       subtitle = "Top 5 player-team-WAR highlighted",
       caption="Source: Baseball Reference")
ggsave(here(paste(ThisLeague, "lol.png")),plot=p_out, height =20, width = 16, units ="in")
```



```{r }
WAR_cum_2teams <- WAR_cum %>% filter (team_ID=="SEA" || team_ID=="LAA")
                                        #team_ID=="TEX"||team_ID=="LAA"||team_ID=="HOU")
WAR_cum_2teams
WAR_cum_2teams[is.na(WAR_cum_2teams)] <- 0
p <- ggplot(WAR_cum_2teams, mapping=aes(x = year_ID, y=cumWAR_by_team)) + geom_line(aes(group=name_common), alpha = .7)+ facet_wrap(~ factor(team_ID))
p + gghighlight(  max(cumWAR_by_team),calculate_per_facet=TRUE, max_highlight = 5)
#+ gghighlight(max(cumWAR_by_team)) + facet_wrap(~ factor(team_ID), ncol=5)
```





